---
title: "Occupy Women: Gender Disparities in the US Film Industry"
author: "Cathy Lee, Dardalie Brooks, and Ha Cao"
date: "20 April 2018"
output: 
  html_document:
    theme: united
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
---

Please see our Github repository[^github] for more details on this project.

## Introduction
In the film industry, women are underrepresented: although they account for 50% of moviegoers[^moviegoers], they make up a startlingly small percentage of producers, writers, directors[^underrepresented], and other significant roles in the films they watch. Our objective was to find out the genres and roles in which gender disparity is the most glaring, and to find out whether women are discriminated against beyond representation: would movies with female-dominated production teams receive less funding than movies with male-dominated production teams?

## Scope
We filtered for movies after 1980 to keep our analysis up-to-date with current events. We wrangled data from movies produced in the US by US companies, so that we could use the US social climate as a control factor. Including movies made in other countries with different outlooks on gender diversity would make the data difficult to analyse. Therefore, we filtered for movies made in USD, located in the USA, and made by US film companies.

## Data Retrieval from SQL Database
From the `IMDB`database, we pulled the movie ID, title, production year, genre, budget, and gross revenue of movies. We pulled another data frame about gender and role of each person in the crew (including actors/actresses). The last data frame we extracted was the production team members (director, producer, writer) and the budget of corresponding movies.

```{r message=FALSE, warning=FALSE}
# set up the environment
library(mdsr)
library(RMySQL)
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(formattable)
theme_set(theme_pubr())
db <- dbConnect_scidb(dbname = "imdb")

# query to get basic information about the movies
# including movie ID, title, production year, genre, budget, and gross revenue in the US
q1 <- "
SELECT t.id, t.title, t.production_year, mii.info AS genre,
mii2.info AS budget, mii3.info AS gross
FROM title t 
JOIN movie_info mii ON mii.movie_id = t.id
JOIN movie_info mii2 ON mii2.movie_id = t.id
JOIN movie_info mii3 ON mii3.movie_id = t.id
JOIN movie_companies mc ON mc.movie_id = t.id
JOIN company_name cn ON cn.id = mc.company_id
WHERE t.kind_id = 1
 AND mii.info_type_id = 3
 AND mii2.info_type_id = 105
 AND mii2.info LIKE '$%'
 AND mii3.info_type_id = 107
 AND cn.country_code = '[us]'
 AND production_year > 1980
 AND mii3.info LIKE '%(USA)'
ORDER BY mii2.info desc;
"
movies_all_info <- db %>%
  dbGetQuery(q1)

# query to get information about crew personnel of the movies
q2 <- "
SELECT ci.movie_id, ci.person_role_id AS char_id, 
n.gender, rt.role 
FROM cast_info ci
JOIN movie_info mii ON mii.movie_id = ci.movie_id
JOIN title t ON mii.movie_id = t.id
JOIN movie_info mii2 ON mii2.movie_id = t.id
JOIN name n ON n.id = ci.person_id
JOIN role_type rt ON rt.id = ci.role_id
JOIN movie_companies mc ON mc.movie_id = t.id
JOIN company_name cn ON cn.id = mc.company_id
WHERE mii.info_type_id = 105
 AND mii.info LIKE '$%'
 AND mii2.info_type_id = 107
 AND cn.country_code = '[us]'
 AND t.production_year > 1980
 AND mii2.info LIKE '%(USA)'
 AND t.kind_id = 1
ORDER BY mii.info DESC;
"
crew_info <- db %>%
  dbGetQuery(q2)

# query to get movies made by US companies only 
# get gender and budget information for each movie
# and info about production team (director, producer, writer)
sql <- "
SELECT t.title, mi.info AS budget, n.gender, t.production_year
FROM cast_info ci
JOIN name n ON n.id = ci.person_id
JOIN movie_info mi ON mi.movie_id = ci.movie_id
JOIN movie_info mi2 ON mi2.movie_id = mi.movie_id
JOIN title t ON t.id = mi.movie_id
JOIN movie_companies mc ON mc.movie_id = t.id
JOIN company_name cn ON cn.id = mc.company_id
WHERE mi.info_type_id = 105
  AND mi.info LIKE '$%'
  AND ci.role_id IN (3,4,5)
  AND mi2.info_type_id = 107
  AND mi2.info LIKE '%(USA)'
  AND cn.country_code = '[us]'
  AND t.production_year > 1980
  AND t.kind_id = 1
ORDER BY t.title; 
"
budget_by_gender <- db %>%
  dbGetQuery(sql)
```

## Data Analysis 

### Gender Representation by Role in the Crew
```{r warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5 }
# a table that has a total number of people as a column
# in order to join with crew_short later
crew_total <- crew_info %>%
  mutate(role_in_movie = ifelse(role %in% c("actor", "actress"), "actor/actress", role)) %>%
  group_by(movie_id, role_in_movie) %>%
  drop_na(gender) %>%
  summarize(total = n()) %>%
  group_by(role_in_movie) %>%
  summarize(total = sum(total))

# specific info about gender distribution within each role of the crew for each movie
crew_short <- crew_info %>%
  mutate(role_in_movie = ifelse(role %in% c("actor", "actress"), "actor/actress", role)) %>%
  group_by(movie_id, role_in_movie, gender) %>%
  summarize(num = n()) %>%
  drop_na(gender) %>%
  group_by(role_in_movie, gender) %>%
  summarize(total_by_gender = sum(num)) %>%
  inner_join(crew_total, by = "role_in_movie") %>%
  mutate(prop = total_by_gender/total) %>%
  group_by(role_in_movie) %>%
  mutate(prop_female = first(prop))

# create a stacked bar chart of gender ratio for different roles in movies
ggplot(data = crew_short, aes(x = reorder(role_in_movie, -prop_female), y = prop, fill = gender)) +
  geom_bar(stat = "identity", position = "fill", alpha = 0.5) +
  scale_y_continuous(name = "Percentage", labels = scales::percent) +
  theme_gdocs() +
  labs(title = "Gender Representation by Role") +
  theme(plot.title = element_text(size = 12),
        legend.title = element_text(size = 10)) +
  scale_fill_manual(values = c("#f8766d", "#00bfc4"), name = "Gender", breaks = c("f", "m"), labels = c("F", "M")) +
  scale_x_discrete(name = "Role in Crew", labels = c("cinematographer" = "Cinematographer", "composer" = "Composer",
                              "director" = "Director", "writer" = "Writer",
                              "editor" = "Editor", "production designer" = "Production Designer",
                              "producer" = "Producer", "miscellaneous crew" = "Miscellaneous Crew", 
                              "actor/actress" = "Actor/Actress", "costume designer" = "Costume Designer")) +
  geom_text(aes(label = percent(prop), y = prop), size = 3, position = position_stack(vjust = 0.5)) +
  coord_flip() 
```
Almost all employment fields are male-dominated except costume designing (almost three-quarters female). Although costume designing plays a pivotal role in movie production, its weekly base salary is $2,789.73, low compared to that of production designers, a similar field that is male-dominated, boasting a weekly base salary of $3,754.47[^salary]. Even in the few fields that are female-dominated, women face gender inequality. 

Over the 38-year period we selected, women accounted for 4% of directors. This statistic was 8% in 2017, indicating that there might be some positive change where gender diversity is concerned. However, the cinematographer field is the most heavily male-dominated - our data showed that women made up only 1.3% of cinematographers in the US film industry. This percentage, unlike that of directors, has barely changed - it only rose slightly to 2% in 2017[^2017stats]. 

Some contend that gender discrimination in the film industry is non-existent - making movies takes a toll on personal relationships, and filming might be on locations far from one’s primary residence. Women are usually the primary caregivers of the family[^caregivers], and cannot afford to be away from home for prolonged periods, and thus do not apply for jobs in the film industry as often as men. However, this is a misconception: interviews conducted by the US Equal Employment Opportunity Commission (EEOC) uncovered hiring discrimination faced by women who applied to work in the film industry, but were rejected because the crew “[didn’t] like working with women”, even though the women applying were sufficiently qualified[^qualified]. Therefore the issue does not lie in a deficiency of women applying for jobs in the film industry, but rather with discrimination in the hiring process.

### Gender Representation by Genre 
```{r warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5 }
# codes for the whole crew in general
# a table that has a total number of people as a column
# in order to join with crew_general later
crew_total_2 <- crew_info %>%
  group_by(movie_id, gender) %>%
  summarize(total = n()) %>%
  drop_na(gender) %>%
  group_by(movie_id) %>%
  summarize(total = sum(total))

# general info about gender distribution (all roles together) for each movie
crew_general <- crew_info %>%
  group_by(movie_id, gender) %>%
  summarize(num = n()) %>%
  drop_na(gender) %>%
  inner_join(crew_total_2, by = "movie_id") %>%
  mutate(prop = num/total)

# codes for actors/actresses specifically
# a table that has a total number of people as a column
# in order to join with crew_general_acting later
crew_total_acting <- crew_info %>%
  filter(role %in% c("actor", "actress")) %>%
  group_by(movie_id, gender) %>%
  summarize(total = n()) %>%
  drop_na(gender) %>%
  group_by(movie_id) %>%
  summarize(total = sum(total))

# general info about gender distribution of actors/actresses for each movie
crew_general_acting <- crew_info %>%
  filter(role %in% c("actor", "actress")) %>%
  group_by(movie_id, gender) %>%
  summarize(num = n()) %>%
  drop_na(gender) %>%
  inner_join(crew_total_acting, by = "movie_id") %>%
  mutate(prop = num/total)

# top 10 most popular genres (most movies listed in that genre)
top_genres <- movies_all_info %>%
  group_by(genre) %>%
  summarize(num_movies = n()) %>%
  arrange(desc(num_movies)) %>%
  head(10)

# filter out top 10 genres
top_genres_info <- movies_all_info %>% 
  filter(genre %in% c("Drama", "Comedy", "Thriller", "Action", "Adventure", "Crime", "Romance", "Sci-Fi", "Fantasy", "Horror")) %>%
  group_by(genre, id) %>%
  summarize(title = first(title),
            production_year = first(production_year),
            budget = first(budget),
            gross = first(gross))

# parse $ budget and gross into real numbers to calculate
top_genres_info <- top_genres_info %>%
  mutate(real_budget = parse_number(budget),
         real_gross = parse_number(gross))

# join crew_general with top_genres_info to get the gender distribution of the whole crew for each movie within each genre
gender_by_genre_crew <- top_genres_info %>%
  inner_join(crew_general, by = c("id" = "movie_id")) %>%
  group_by(genre, gender) %>%
  summarize(budget = mean(real_budget),
            gross_revenue = mean(real_gross),
            prop = mean(prop)) %>%
  group_by(genre) %>%
  mutate(prop_female = first(prop))

gender_by_genre_acting <- top_genres_info %>%
  inner_join(crew_general_acting, by = c("id" = "movie_id")) %>%
  group_by(genre, gender) %>%
  summarize(budget = mean(real_budget),
            gross_revenue = mean(real_gross),
            prop = mean(prop)) %>%
  group_by(genre) %>%
  mutate(prop_female = first(prop))

# graphs about gender distribution by genre
# create a stacked bar chart for different roles in movies
gender_crew_plot <- ggplot(data = gender_by_genre_crew, aes(x = reorder(genre, -prop_female), y = prop, fill = gender)) +
  geom_bar(stat = "identity", position = "fill", alpha = 0.5) +
  scale_x_discrete(name = "Genre") +
  scale_y_continuous(name = "Percentage", labels = scales::percent) +
  theme_gdocs() +
  labs(title = "Gender Representation in Whole Crew \nby Genre") +
  theme(plot.title = element_text(size = 12),
        legend.title = element_text(size = 10)) +
  scale_fill_manual(values = c("#f8766d", "#00bfc4"), name = "Gender", breaks = c("f", "m"), labels = c("F", "M")) +
  geom_text(aes(label = percent(prop), y = prop), size = 3, position = position_stack(vjust = 0.5)) +
  coord_flip()

gender_acting_plot <- ggplot(data = gender_by_genre_acting, aes(x = reorder(genre, -prop_female), y = prop, fill = gender)) +
  geom_bar(stat = "identity", position = "fill", alpha = 0.5) +
  scale_x_discrete(name = "Genre") +
  scale_y_continuous(name = "Percentage", labels = scales::percent) + 
  theme_gdocs() +
  labs(title = "Gender Representation in Cast Only \nby Genre") +
  theme(plot.title = element_text(size = 12),
        legend.title = element_text(size = 10)) +
  scale_fill_manual(values = c("#f8766d", "#00bfc4"), name = "Gender", breaks = c("f", "m"), labels = c("F", "M")) +
  geom_text(aes(label = percent(prop), y = prop), size = 3, position = position_stack(vjust = 0.5)) +
  coord_flip()

figure <- ggarrange(gender_crew_plot, gender_acting_plot,
                    ncol = 2, nrow = 1)
figure
```

```{r warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5 }
gross_plot <- ggplot(data = gender_by_genre_crew, aes(x = reorder(genre, -gross_revenue), y = gross_revenue)) +
  geom_bar(stat = "identity", alpha = 0.5, fill = "#a081f4") + 
  theme_gdocs() +
  theme(axis.text.x = element_text(angle = 45)) +
  scale_x_discrete(name = "Genre") +
  scale_y_continuous(name = "Gross Revenue (USD)", labels = scales::comma) +
  labs(title = "Average Gross Revenue per Movie for Top 10 Genres") +
  theme(plot.title = element_text(size = 12))
gross_plot
```
We also investigated how gender distribution of the cast (actors and actresses) and of the entire crew (all roles) might differ according to various film genres, then compared these percentages to the gross revenues by genre. The "Adventure" genre, the second most male-dominated genre by cast (76.66% male) and crew (78.21% male), is also the highest grossing genre. Some might use this as justification for the lack of gender parity in the industry, since a cast and crew does not require gender equality to reap profits. Nevertheless, the graphs show that fantasy has the third highest average gross revenue per movie, and is the fourth most gender-diverse genre; correlation does not equal causation, and factors other than gender representation influence a movie’s revenue. Overall, the lack of women in all genres could be attributed to male-led movies being considered universally relatable and more complex, while female-led movies are othered and trivialised[^othered].

### Change of Gender Representation in the Whole Crew by Genre from 1981-2017
```{r warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5 }
# all top 10 genres_crew ratios_over time 
gender_crew_time_top10 <- top_genres_info %>% 
  inner_join(crew_general, by = c("id" = "movie_id")) %>%
  group_by(genre, gender, production_year) %>%
  summarize(prop_female = mean(prop)) 
time_f_top10 <- subset(gender_crew_time_top10, gender_crew_time_top10$gender == "f")

# plot All genres
timeplot <- ggplot(data = time_f_top10,
                      aes(x = production_year, y = prop_female)) +
  geom_line(col = "#8f42f4", size = 1) +
  scale_x_continuous(name = "Year")+
  scale_y_continuous(name= "Percentage of Female in Whole Crew", labels = scales::percent)+
  theme_gdocs() +
  facet_wrap(~genre) +
  labs(title = "Change in Gender Ratio by Genre from 1981-2017") +
  theme(plot.title = element_text(size = 12))

timeplot

# gender ratio's overall change over the last 36 years for each genre
top_genre_percent_change <- time_f_top10 %>%
  filter(production_year %in% c("1981", "2017")) %>% 
  group_by(genre) %>% 
  summarize(prop_change = last(prop_female) - first(prop_female)) %>%
  arrange(desc(prop_change))

top_genre_percent_change
# the mean change of gender ratio across all genres
mean_change <- mean(top_genre_percent_change$prop_change)
mean_change
```
We observed how the percentage of female crew members changed from 1981 to 2017 in the top ten revenue-generating genres. The “Romance” genre has the biggest change in gender ratio (28.2%) over the last 36 years, and is the most gender diverse, with “Comedy” coming in second. However, romantic comedies have dialogue that is approximately 58% male, even those with female leads[^dialogue]. The relatively high female participation in “Romance” can be attributed to the prevalence of heterosexuality in mainstream media, requiring a man and woman - in 2015, only 17.5% of films contained LGBT characters, and 65% of those characters were gay men, compared to 10% lesbians. Dishearteningly, there might be more women in comedies because they are often at the root of sexist jokes[^comedies]. 

### Budget Gap between Production Teams that are Different in Majority Gender
```{r warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5 }
# calculate the gender demographics on the production team of each movie
# calculate whether a movie's production team is female or male dominated 
budget_majority_gender <- budget_by_gender %>%
  mutate(parsed_budget = parse_number(budget)) %>%
  group_by(title) %>%
  drop_na(gender) %>%
  summarize(production_year = first(production_year),
            num_male = sum(gender == 'm'), 
            num_female = sum(gender == 'f'),
            parsed_budget = mean(parsed_budget)) %>%
  mutate(dominant_gender = ifelse(num_male >= num_female, ifelse(num_male > num_female, "majority_male", "equal"), "majority_female")) %>%
  arrange(desc(parsed_budget)) 

# plot the boxplot
budget_gender_plot <- budget_majority_gender %>%
  ggplot(aes(x = as.factor(dominant_gender), y = parsed_budget, fill = dominant_gender)) +
  geom_boxplot(alpha = 0.5) +
  theme_gdocs() +
  ggtitle("Majority Gender in a Film's Production Team vs. Budget Received") +
  scale_x_discrete(name = "Majority Gender in Production Team",
                   limits = c("majority_male", "equal", "majority_female"),
                   labels = c("Majority male", "Equally represented", "Majority female")) +
  scale_fill_manual(values = c( "#a081f4","#f8766d","#00bfc4")) +
  scale_y_continuous(name = "Budget received (USD)", 
                     labels = scales::comma,
                     breaks = seq(0, 300000000, 50000000)) +
  guides(fill = FALSE)
budget_gender_plot
```

We investigated the relationship between the majority gender in a film’s production team (comprising writers, producers, and directors) and the budget that film received. Interestingly, female-dominated teams had a lower upper quartile *and* lower quartile than male-dominated teams. Male-dominated teams also have far more outliers compared to 4 each for the other 2 teams, indicating that there were many movies with exorbitant budgets given to male-dominated teams. Although the median is roughly the same for male and female-dominated teams, indicating that male-dominated teams are often given lower budgets than higher, the fact remains that the highest budgets are largely given to male-dominated teams.

Upon further research, we found that film industry decision-makers perceive that female directors make films tailored for only a subset of the market, while male directors are expected to make films that are more generalised and wide-reaching. They also doubt the competence of female directors, and wrongly assume a scarcity of female directors (when in fact they are simply unaware of them)[^decisionmakers]. This could account for why men are more likely to be hired over women for production teams.

```{r warning = FALSE, MESSAGE = FALSE}
# run a regression model to test whether the difference in budget between male and female-dominated production team is significant
budget_majority_gender$dominant_gender <- as.factor(budget_majority_gender$dominant_gender)
budget_majority_gender$dominant_gender <- relevel(budget_majority_gender$dominant_gender, ref = "majority_male")
budget_model <- lm(data = budget_majority_gender, parsed_budget ~ dominant_gender + production_year)
summary(budget_model)
confint(budget_model, level = .95) # obtain 95% CIs for all coefficients
```
We performed multiple regression to analyse the differences in budget received among the 3 categories. `95% CI of the coefficient of dominant_gendermajority_female = [-23348725, -4892295]`, meaning that we are 95% confident that female-dominated production teams have from 23.35 to 48.92 million USD on average less in budget than male-dominated teams per movie. 

The coefficient of -13.04 million indicates that production teams with equal gender representation are predicted to have (on average) 13.04 million USD less in budget than male-dominated teams, adjusted for production year. This coefficient is statistically significant because our p-value is 0.007 (< 0.05); there is sufficient evidence that this budget gap is statistically significant.

The coefficient of -14.12 million indicates that female-dominated production teams are predicted to have an average budget $14.12m less than male-dominated production teams, adjusted for production year. This coefficient is statistically significant because our p-value is 0.003 (< 0.05); there is sufficient evidence that this budget gap is statistically significant.

[^github]: Our [Github Repository](https://github.com/cathyleee/sds192-mp4)
[^moviegoers]: Statistic from [MPAA 2017](https://www.mpaa.org/wp-content/uploads/2018/04/MPAA-THEME-Report-2017_Final.pdf)
[^underrepresented]: Women comprised only 24% of protagonists in 2017 according to the [Center for the Study of Women in Television and Film](https://womenandhollywood.com/resources/statistics/), and were in the extreme minority where significant roles in production were concerned.
[^salary]: The salaries of the Art Directors Guild, whose members are 73% male, and that of the Costume Designers Guild, whose members are almost 80% female, were compared in a [Variety article](http://variety.com/2016/biz/news/art-director-costumes-groups-gender-inequality-1201905400/).
[^2017stats]: Retrieved from a study conducted by the [Center for the Study of Women in Television and Film](https://womenintvfilm.sdsu.edu/wp-content/uploads/2018/01/2017_Celluloid_Ceiling_Report.pdf)
[^caregivers]: Women are found by the [Pew Research Centre](http://www.pewresearch.org/fact-tank/2015/10/01/women-more-than-men-adjust-their-careers-for-family-life/) to be more likely than men to adjust their careers for family life.
[^qualified]: From excerpts of interviews on gender discrimination in the film industry conducted by the [EEOC](http://www.scpr.org/programs/the-frame/2017/02/16/55132/gender-discrimination-in-hollywood-is-real/).
[^othered]: We are subconsciously more critical of female-led movies, and more likely to see women as two-dimensional. Retrieved from [the Guardian](https://www.theguardian.com/news/2018/mar/06/the-male-glance-how-we-fail-to-take-womens-stories-seriously).
[^dialogue]: This statistic can be attributed to the greater number of male supporting characters in romantic comedies, even in those with female leads. Retrieved from [a study on film dialogue by The Pudding](https://pudding.cool/2017/03/film-dialogue/).
[^comedies]: Comedies are much less likely to be palatable for male audiences if humour is seen from a woman's perspective. Retrieved from [IndieWire](http://www.indiewire.com/2011/04/the-misogyny-machine-that-rules-hollywood-comedies-212615/).
[^decisionmakers]: Obstacles female directors face in the film industry by the [Female Filmmakers Initiative](http://www.sundance.org/pdf/artist-programs/wfi/phase-iii-research---female-filmmakers-initiative.pdf).
